---
title: Generating synthetic data with R
subtitle: ""
---



```{r, include=FALSE, cache=FALSE}
#R script that contains code chunks shown below
knitr::read_chunk('../../../code/synthetic-data-new-r-example-1-code.R')
knitr::read_chunk('../../../code/synthetic-data-new-r-example-2-code.R')
```


# Overview
In this unit, we look at a few examples of generating synthetic data with `R.


# Learning Objectives

* Be able to generate synthetic data with `R` code.


# Introduction

While there are `R` packages available that help with the generation of synthetic data - and we will discuss them in a different unit, it is also fairly easy to use basic `R` functionality to generate simulated data. We'll do a few examples to show how this can be done.

AI tools are very useful for generating synthetic data. In fact, the code in the examples below was partially generated through back and forth with ChatGPT, together with some manual editing. More on using AI to generate data is provided in another unit of this module, and the more general topic of LLM AI tool use for coding is covered in the [AI module](../../module-ai/ai-overview.qmd).


# Example 1

For the first example, we'll generate a simple dataset of some hypothetical clinical trial for some treatment (say, a Cholesterol lowering medication) and its impact on outcomes such as Cholesterol levels and adverse events.

:::note
It is often useful to go slow and type your own code, or copy and paste the code chunks below and execute one at a time. However, if you are in a hurry, you can find the code for this example in [this file](/code/synthetic-data-new-r-example-1-code.R).
:::

:::note
In the code below, you might be wondering what the `dplyr::glimpse()` and similar such notation is about. This notation indicates explicitly from which package the function comes. Once you loaded a package, that's not necessary anymore, just using `glimpse()` would work equally. But it is often nice to quickly see what package a function comes from. An important thing to note is that some packages share function names. In such cases, if a function appears in multiple packages in your environment, `R` will default to using the one from the most recently loaded package. Using explicit notation is a good way to be sure you're using the function you want! Thus, you'll see me use the explicit notation often, though not always. It's typically a matter of being more explicit versus typing less. I switch back and forth.
:::


## Setup

We'll start by loading the packages used in the code below. While we will not use packages that are specifically meant to generate synthetic data, we will still use several common packages that make data generation tasks easier. That said, you could also do all of this with base `R` and no additional packages.


```{r, packages, message=FALSE, warning=FALSE}
```

```{r, setup, message=FALSE}
```


## Generating data 

```{r, makedata}
```


## Checking data 

Take a peek at the generated data.


```{r, checkdata}
```


This concludes our first example. This was a very simple setup, with rectangular data. You often encounter these kind of data in clinical trials. As you generate your data, you can build in any dependencies between variables you want to explore. Then later, when you use the data to test your analysis code, you can see if your analysis code can detect the dependencies you built in. We'll come back to that. 

You could also add further complexities into your synthetic data, for instance you could set some values to be missing, or you could add some outliers. The goal is to generate data that has as the important features of your real dataset to allow you to test your analysis approach on data where you now the truth (since you generated it). If your analysis works on your generated data, there is hope it might also work on the real data (for which of course you don't know the truth). We'll define in a later unit what we mean by "your analysis works". But basically, you want to be able to recover the patterns/dependencies you built into your data with your analysis methods.


# Example 2

In this example, we'll generate a dataset that has a more complex structure. We'll generate a dataset that has a longitudinal structure, i.e. repeated measurements over time. We'll also generate a dataset that has a hierarchical structure, i.e. measurements nested within groups.    

We'll walk slowly through the code junks, you can find all the code in one file [here](code/synthetic-data-new-r-example-2-code.R).



## Setup

The usual setup steps.

```{r, packages-ex2, message=FALSE, warning=FALSE}
```

```{r, setup-ex2, message=FALSE}
```


## Generating data 

```{r, makedata-ex2}
```


## Checking data 

Take a peek at the generated data.


```{r, checkdata-ex2}
```


## Exploratory plot

```{r, plotdata-ex2}
```

This dataset has a bit more complicated structure than the previous one, but it isn't much harder to generate it with code. You'll want to generate your data such that it mimics the real data you plan on analyzing.

## Save data

```{r, savedata-ex2}
```

We are saving the data as CSV and Rds files. CSV files are readable with pretty much any software and thus very portable. Rds files are `R`-specific, thus not as flexible. The advantage of Rds files is that they are generally smaller, and they retain information about the variables, e.g. if they are factor or numeric variables. Either format works, sometimes one or the other might be better.



# Summary

It is fairly easy to generate synthetic data with some basic `R` commands, and it's even easier if you use some LLM AI tool to help you write the code. 

Since you generate the data, you can decide what dependencies between variables you want to add, and you can then test your models to see if they can find those patterns. We'll get to that in another unit.


# Further Resources

While the manual generation of data is generally fairly easy and gives you the most control, sometimes using `R` packages specifically designed for this purpose can be useful. We'll look at a few in a later unit. 


